Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility

### 2025-11-17 paper/paper.tex
up v2
Big up methods and results

% Colores extra por nombre
\usepackage[dvipsnames]{xcolor}
% Ajustes de hipervínculos
% \hypersetup{
%   colorlinks = true,
%   citecolor  = DarkSlateBlue,     % << citas
%   urlcolor   = DarkSlateBlue  % URLs/DOIs
% }
% 
\begin{document}
\maketitle
\begin{abstract}
My abstract \newline \textbf{Keywords}: pension commodification, social
mobility, causal inference, Chile, public preferences
\end{abstract}

\setstretch{1.5}
This document was last modified at 2025-11-03 17:13:25 and it was last
rendered at 2025-11-03 17:13:25.

@@

Chile offers a compelling case for examining individual preferences for
market justice in the access to core social services. Since the early
1990s, the country experienced a long period of sustained economic
growth, with per capita GDP increasing by nearly 4\% annually until the
mid-2010s (\citeproc{ref-llorca-jana_historia_2021}{Llorca-Jaña \&
Miller, 2021}). This growth brought notable improvements in living
standards, including a strong decline in income poverty---from 38.6\% in
1990 to 6.5\% in 2022 (\citeproc{ref-mideplan_evolucion_2017}{MIDEPLAN,
2017}). However, this success story has been tempered by growing
concerns over the unequal distribution of its benefits. In recent years,
economic growth has slowed---exacerbated by the global impact of the
COVID-19 pandemic---and structural inequalities have become more visible
(\citeproc{ref-barozet_clases_2021}{Barozet, Contreras, et al., 2021}).
Despite its economic achievements, Chile remains one of the most unequal
countries in Latin America and the OCDE. The poorest 50\% of the
population captures just 10\% of total income and holds negative net
wealth, while the wealthiest 1\% receives nearly 27\% of all income and
controls almost half (49.6\%) of the country's wealth
(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}), a figure that
has seen little change since the 1990s
(\citeproc{ref-flores_top_2020}{Flores et al., 2020}). These stark
disparities extend beyond income, reflecting deep inequalities in access
to essential social services (\citeproc{ref-pnud_desiguales_2017}{PNUD,
2017}).

@@

Over recent decades, Chile has undergone a significant transformation in
its class structure, shaped by sustained economic growth, the expansion
of the service sector, the massification of education, and major changes
in the role of the state and social policy
(\citeproc{ref-ruiz_chilenos_2014}{Ruiz \& Boccardo, 2014}). These
changes contributed to the growth of the middle sectors, a phenomena
labelled as ``mesocratrization'' of society
(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013}).
However, this expansion did not reduce structural inequalities nor
facilitate upward mobility into elite positions
(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}). While Chile shows high absolute mobility, most flows occur
between adjacent classes---particularly from working classes into lower
service-class positions---while access to upper classes remains limited,
indicating strong elite closure and reproduction
(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014};
\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}; \citeproc{ref-perez-ahumada_clases_2019}{Pérez-Ahumada, 2019};
\citeproc{ref-torche_unequal_2005}{Torche, 2005}). Though often
described as ``unequal but fluid'', recent evidence points to growing
rigidity in relative mobility, especially after the economic slowdown,
increased household debt, labor market precarity, and the COVID-19
crisis (\citeproc{ref-barozet_clases_2021}{Barozet, Contreras, et al.,
2021}). However, the majority of the population identifies themselves as
members of the middle class (\citeproc{ref-castillo_todos_2013}{Castillo
et al., 2013}). These dynamics have fueled a legitimacy crisis:
meritocracy is increasingly questioned, and many perceive that
advancement depends more on informal networks than on effort or earned
credentials (\citeproc{ref-barozet_clases_2021}{Barozet, Contreras, et
al., 2021}). The combination of constrained mobility, persistent
inequality, and concentrated opportunity has eroded expectations of
upward mobility and intensified social discontent
(\citeproc{ref-pnud_desiguales_2017}{PNUD, 2017}).

@@

Social mobility effects---understood as impacts on an outcome arising
from movements between an origin state and a destination state (e.g.,
social class)---have long been a focus of sociological research
(\citeproc{ref-eyles_social_2022}{Eyles et al., 2022};
\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}).
Yet, as Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024, p.
467}) emphasize, most of the mobility hypotheses are, at their core,
individual-level counterfactual comparisons between the observed outcome
under a mobility trajectory and the outcome the same person would have
shown if they had remained in their origin class (or moved to an
alternative destination). Standard specifications (e.g., SAM, DRM, and
mobility-contrast models) struggle to retrieve these inherently
counterfactual quantities due to identification constraints and their
reliance on between-group contrasts constructed from additive terms and
interactions, thereby yielding primarily associational evidence
(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024};
\citeproc{ref-song_there_2025}{Song \& Zhou, 2025}). I therefore adopt
Breen and Ermisch's (\citeproc{ref-breen_effects_2024}{2024}) causal
formulation, treating destination class as the treatment, conditioning
on the class of origin, and estimating heterogeneous destination effects
with observational data under explicit identification assumptions. To
align design and target, I follow the MIDA template
(\citeproc{ref-blair_research_2023}{Blair et al., 2023}): set out the
causal model and assumptions (M), define the inquiry and estimand (I),
describe the data and variables (D), and detail the answer strategy (A)
used to identify the causal effect of intergenerational social mobility
on preferences for the commodification of pensions.

@@

Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}), building on
a critical reassessment of the inferential limits of standard mobility
models, propose a causal framework that reconceptualizes the ``mobility
effect'' as the treatment effect of reaching a destination class, with
impacts heterogeneous by class of origin. The core claim is that typical
mobility hypotheses are fundamentally within-person counterfactual
comparisons rather than between-person contrasts. Defining mobility as
\(M = D - O\) ---a change from origin \((O)\) to destination \((D)\),
rather than an additive or interactive combination---places the focus
squarely on the Neyman--Rubin problem: for any individual, we observe
the result in the social position they currently occupy, but not in the
situation of other alternative destinations or immobility.

@@

In fact, establishing the causal relationship between objective social
mobility and subjective outcomes, such as preferences, poses several
challenges for causal inference. Below, I use directed acyclic graphs
(DAGs) to illustrate some of these challenges and evaluate possible
strategies for identifying the effects of social mobility.
Figure~\ref{fig-dag} depicts the causal model guiding identification. In
this model, a person's preference (\(Y\)) is directly caused by her
class destination (\(D\)) and can also be affected by different
attributes determined in childhood or earlier (\(C\)), such as family
resources, household composition, and ascriptive status (pre-treatment
confounders). Class origin (\(O\)) is determined by those
origin-attributes (\(C\)) and, in turn, affects destination (\(D\)). In
addition, unobserved factors (\(U\)) can also influence
origin-attributes (\(C\)). There is no direct arrow from \(O\) to \(Y\),
reflecting the assumption that the effect of origin on the outcome
operates entirely through background factors \((C)\) and destination
\((D)\). Accordingly, conditioning on \((O,C)\) could blocks the
relevant back-door paths from \(D\) to \(Y\), specifically
\(D\leftarrow O\leftarrow C\rightarrow Y\) and, if present,
\(D \leftarrow O \rightarrow Y\), so \((O,C)\) is a minimal sufficient
adjustment set.

@@

Formally, the DAG encodes the following relations:

\[
U \rightarrow C,\quad C \rightarrow O,\quad C \rightarrow D,\quad C \rightarrow Y,\quad O \rightarrow D,\quad D \rightarrow Y
\]

This structure corresponds to a nonparametric structural causal model in
which \(D\) is the treatment variable, \(Y\) the outcome, and \((O, C)\)
the minimal sufficient adjustment set. Conditioning on these variables
satisfies the back-door criterion for identifying the causal effect of
mobility on preferences (\citeproc{ref-breen_effects_2024}{Breen \&
Ermisch, 2024}).

Why are the assumptions plausible here? Conditional independence is
targeted by controlling for a plausible sufficient adjustment set
\((O,C)\); inverse probability weights (IPW) obtained via entropy
balancing enforces covariate balance within each origin stratum.
Positivity is assessed via propensity scores and trimming of extreme
weights, restricting to common support. SUTVA holds by treating the
destination class as a single exposure measured before \((Y)\) and
assuming no interference. Together, the DAG and assumptions define
conditions for identifying the causal effect of intergenerational social
mobility on preferences for commodification.

@@

The causal inquiry guiding this study asks: How does intergenerational
social mobility affect individuals' preferences for the commodification
of pensions in Chile? Following the potential outcomes framework
proposed by Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024,
p. 473}), the estimand of interest is the average treatment effect on
the treated (ATT), defined as the mean causal effect of moving from an
origin class \((O=j)\) to a destination class \((D=k)\), compared with
the counterfactual outcome that those same individuals would have
exhibited had they instead reached an alternative destination \((k')\):

\begin{equation}\phantomsection\label{eq-breen}{ 
ATT_{j,k,k′} =E[Y(D=k|O=j, D=k)]−E[Y(D=k′|O=j, D=k)]
}\end{equation}

In this study, I focus on the specific case where the counterfactual
destination corresponds to class immobility \((k' = j)\), that is, how
the outcome among people from origin \emph{j} who entered destination
\emph{k} would have bee different if those people had, counterfactually,
remained in origin \emph{j} instead (inmobility). Accordingly, the
estimand becomes:

\begin{equation}\phantomsection\label{eq-att}{ 
ATT_{j,k,j} =E[Y(D=k|O=j, D=k)]−E[Y(D=j|O=j, D=k)]
}\end{equation}

which represents the average causal effect of attaining class
\emph{(k)}---for instance, moving upward or downward---relative to
remaining in class \emph{(j)}, among individuals who actually
experienced such mobility. This estimand captures the within-origin
counterfactual contrast at the heart of mobility effects: how a person's
preference for market-based pensions would differ if, counterfactually,
they had stayed in their origin class rather than moved to their
observed destination.

@@

Because respondent occupation is not measured in every wave, I restrict
the analysis to the 2016, 2018, and 2023 waves. Parental occupation was
collected only in 2023 as open-text labels. I coded these texts to the
ISCO-08 two-digit scheme using the National Institute of Statistics of
Chile's automated coding API and retained cases with high-confidence
matches (\textgreater95\%). Treating parental occupation as a
time-invariant origin attribute, I carried these codes back to 2016 and
2018. After listwise deletion and restricting to key variables
(respondent occupation and the outcome measure), the final analytic
sample comprises \emph{N} = 3,435 observations nested within \emph{N} =
1,787 individuals (2016: 914; 2018: 1,377; 2023: 1,144). Consistent with
the study design, estimation proceeds within trajectory-specific
subsamples (e.g., low→low; low→middle), retaining movers and matched
non-movers from the same origin class. Then, I apply weights to
construct the corresponding counterfactual control sample for each
trajectory. Following Breen and Ermisch
(\citeproc{ref-breen_effects_2024}{2024}), I use the three waves in the
estimations that follow, allowing for correlation between the
observations for each individual in calculating the standard errors
(i.e.~clustering on the personal identifier).

@@

The outcome variable measures preferences regarding the commodification
of pensions, operationalized with a single item addressing how strongly
individuals justify conditioning access to old-age pension benefits
based on individual income. Respondents were asked: ``Is it fair in
Chile that people with higher incomes have better pensions than people
with lower incomes?'' and were asked to indicate their level of
agreement on a five-point Likert scale ranging from 1 (``strongly
disagree'') to 5 (``strongly agree''). For estimation and
interpretability, I dichotomized the item so that values 4-5 indicate
agreement with market-based access to old-age pensions (coded as 1),
while values 1-3 indicate disagreement (coded as 0). This binary coding
treats neutrality as substantially closer to non-support than to support
and produces a clear contrast between respondents who support
market-based differentiation in pensions and those who do not. I use
this item for two main reasons: on the one hand, to be able to make
comparisons with existing literature
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024}) and, on the
other hand, because it captures two fundamental dimensions of
market-based welfare distribution: the role of economic resources as a
key determinant of outcomes and the framing of services as as tradable
commodities that can be bought and sold based on ability to pay
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

Table~\ref{tbl-summary1} shows the descriptive statistics of the
variables used in the study.

>>>>

% Colores extra por nombre
\usepackage[dvipsnames]{xcolor}
% Ajustes de hipervínculos
% \hypersetup{
%   colorlinks = true,
%   citecolor  = DarkSlateBlue,     % << citas
%   urlcolor   = DarkSlateBlue  % URLs/DOIs
% }
% 
\begin{document}
\maketitle
\begin{abstract}
My abstract \newline \textbf{Keywords}: Pension commodification ·
Intergenerational mobility · Causal inference · Meritocracy · Chile
\end{abstract}

\setstretch{1.5}
This document was last modified at 2025-11-17 19:08:04 and it was last
rendered at 2025-11-17 19:08:04.

@@

Chile offers a compelling case for examining individual preferences for
market justice in the access to core social services. Since the early
1990s, the country experienced a long period of sustained economic
growth, with per capita GDP increasing by nearly 4\% annually until the
mid-2010s (\citeproc{ref-llorca-jana_historia_2021}{Llorca-Jaña \&
Miller, 2021}). This growth brought notable improvements in living
standards, including a strong decline in income poverty---from 38.6\% in
1990 to 6.5\% in 2022 (\citeproc{ref-mideplan_evolucion_2017}{MIDEPLAN,
2017}). However, this success story has been tempered by growing
concerns over the unequal distribution of its benefits. In recent years,
economic growth has slowed---exacerbated by the global impact of the
COVID-19 pandemic---and structural inequalities have become more visible
(\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}). Despite its
economic achievements, Chile remains one of the most unequal countries
in Latin America and the OCDE. The poorest 50\% of the population
captures just 10\% of total income and holds negative net wealth, while
the wealthiest 1\% receives nearly 27\% of all income and controls
almost half (49.6\%) of the country's wealth
(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}), a figure that
has seen little change since the 1990s
(\citeproc{ref-flores_top_2020}{Flores et al., 2020}). These stark
disparities extend beyond income, reflecting deep inequalities in access
to essential social services (\citeproc{ref-pnud_desiguales_2017}{PNUD,
2017}).

@@

Over recent decades, Chile has undergone a significant transformation in
its class structure, shaped by sustained economic growth, the expansion
of the service sector, the massification of education, and major changes
in the role of the state and social policy
(\citeproc{ref-ruiz_chilenos_2014}{Ruiz \& Boccardo, 2014}). These
changes contributed to the growth of the middle sectors, a phenomena
labelled as ``mesocratrization'' of society
(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013}).
However, this expansion did not reduce structural inequalities nor
facilitate upward mobility into elite positions
(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}). While Chile shows high absolute mobility, most flows occur
between adjacent classes---particularly from working classes into lower
service-class positions---while access to upper classes remains limited,
indicating strong elite closure and reproduction
(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014};
\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}; \citeproc{ref-perez-ahumada_clases_2019}{Pérez-Ahumada, 2019};
\citeproc{ref-torche_unequal_2005}{Torche, 2005}). Though often
described as ``unequal but fluid'', recent evidence points to growing
rigidity in relative mobility, especially after the economic slowdown,
increased household debt, labor market precarity, and the COVID-19
crisis (\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}).
However, the majority of the population identifies themselves as members
of the middle class (\citeproc{ref-castillo_todos_2013}{Castillo et al.,
2013}). These dynamics have fueled a legitimacy crisis: meritocracy is
increasingly questioned, and many perceive that advancement depends more
on informal networks than on effort or earned credentials
(\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}). The
combination of constrained mobility, persistent inequality, and
concentrated opportunity has eroded expectations of upward mobility and
intensified social discontent (\citeproc{ref-pnud_desiguales_2017}{PNUD,
2017}).

@@

Social mobility effects---understood as impacts on an outcome arising
from movements between an origin state and a destination state (e.g.,
social class)---have long been a focus of sociological research
(\citeproc{ref-eyles_social_2022}{Eyles et al., 2022};
\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}).
Yet, as Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024, p.
467}) emphasize, most of the mobility hypotheses are, at their core,
individual-level counterfactual comparisons between the observed outcome
under a mobility trajectory and the outcome the same person would have
shown if they had remained in their origin class (or moved to an
alternative destination). Standard specifications (e.g., SAM, DRM, and
mobility-contrast models) struggle to retrieve these inherently
counterfactual quantities due to identification constraints and their
reliance on between-group contrasts constructed from additive terms and
interactions, thereby yielding primarily associational evidence
(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024};
\citeproc{ref-song_there_2025}{Song \& Zhou, 2025}). Following the
causal framework of Breen and Ermisch
(\citeproc{ref-breen_effects_2024}{2024}), I conceive of the destination
class as a treatment, conditioning on the origin class and estimating
the heterogeneous effects of the destination with observational data
under explicit identification assumptions. To align design and target, I
follow the MIDA template (\citeproc{ref-blair_research_2023}{Blair et
al., 2023}): set out the causal model and assumptions (M), define the
inquiry and estimand (I), describe the data and variables (D), and
detail the answer strategy (A) used to identify the causal effect of
intergenerational occupational mobility on preferences for the
commodification of pensions.

@@

Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}), building on
a critical reassessment of the inferential limits of standard mobility
models, propose a causal framework that reconceptualizes the ``mobility
effect'' as the treatment effect of reaching a destination class, with
impacts heterogeneous by class of origin. The core claim is that typical
mobility hypotheses are fundamentally within-person counterfactual
comparisons rather than between-person contrasts. Defining mobility as
\(M = D - O\); a change from origin \((O)\) to destination \((D)\),
rather than an additive or interactive combination, places the focus
squarely on the Neyman--Rubin problem: for any individual, we observe
the result in the social position they currently occupy, but not in the
situation of other alternative destinations or immobility.

@@

Establishing the causal relationship between objective social mobility
and subjective outcomes, such as preferences, poses several challenges
for causal inference. Below, I use directed acyclic graphs (DAGs) to
illustrate some of these challenges and evaluate possible strategies for
identifying the effects of social mobility. Figure~\ref{fig-dag} depicts
the causal model guiding identification. In this model, a respondent's
preference \(Y\) is directly affected by her class destination \(D\),
and indirectly influenced by a set of pre-treatment attributes \(C\)
rooted in childhood---family resources, household composition, parental
education, and other ascriptive characteristics. Class origin \(O\) is
itself shaped by these background factors \(C\) and, in turn, affects
destination \(D\). I allow for unobserved determinants \(U\) that
influence background factors \(C\), but I assume that any such
unobserved variation operates only through \(C\). Thus, there are no
direct paths \(U \rightarrow D\) or \(U \rightarrow Y\) beyond those
mediated by \(C\).

Under this structure, conditioning on \((O, C)\) blocks all relevant
backdoor paths from \(D\) to \(Y\)---notably
\(D \leftarrow O \leftarrow C \rightarrow Y\). This renders \((O, C)\) a
minimal sufficient adjustment set for identifying the causal effect of
mobility on preferences, in line with the framework proposed by Breen \&
Ermisch (\citeproc{ref-breen_effects_2024}{2024}). Identification
therefore relies on the assumption that background factors \(C\)
adequately summarize pre-treatment characteristics that jointly shape
both mobility and attitudes.

@@

Why are the assumptions plausible here? Conditional independence is
targeted by controlling for a plausible sufficient adjustment set
\((O,C)\) and by using inverse probability weights obtained via entropy
balancing (\citeproc{ref-hainmueller_entropy_2012}{Hainmueller, 2012}),
which enforce covariate balance within each origin stratum and mitigate
selection on observables. Positivity is assessed by inspecting the
distribution of the entropy-balancing weights and trimming observations
with extreme weights, thus restricting inference to regions of common
support. SUTVA holds by treating the destination class as a single
exposure measured before \((Y)\) and assuming no interference. Together,
the DAG and assumptions define conditions for identifying the causal
effect of intergenerational social mobility on preferences for
commodification.

@@

The causal inquiry guiding this study asks: How does intergenerational
social mobility affect individuals' preferences for the commodification
of pensions in Chile? Substantively, I am interested in the effect of
mobility relative to immobility within a given class of origin.

Following the potential outcomes framework proposed by Breen and Ermisch
(\citeproc{ref-breen_effects_2024}{2024, p. 473}), the estimand of
interest is an average treatment effect on the treated (ATT) defined
within each origin class. I focus on the specific case in which the
counterfactual destination corresponds to class immobility \((k' = j)\).
In other words, I ask how the outcome among people from origin \emph{j}
who entered destination \emph{k} would have bee different if those
people had, counterfactually, remained in origin \emph{j} instead
(inmobility).

Formally, for individuals with origin \((O=j)\) who attain destination
\((D=k)\), the estimand is:

\$\$

ATT\_\{j,k,j\} = E\big[Y(D = k) \mid O = j, D = k \big] -
E\big[Y(D = j) \mid O = j, D = k \big]

\$\$ \{\#eq-att\}

which represents the mean causal effect of moving from origin class
\((j)\) to destination class \((k)\), compared to the counterfactual
outcome those same individuals would have exhibited had they instead
remained in \((j)\). Thus, this ATT captures the counterfactual contrast
within origin that lies at the heart of mobility effects: how
preferences for market-based pensions would differ if, for the same
upwardly or downwardly mobile individuals, their observed destination
class were replaced by immobility in their origin class.

@@

Because respondent occupation\footnote{Parental occupation was collected
  only in 2023 as open-text labels. I coded these texts to the ISCO-08
  two-digit scheme using the National Institute of Statistics of Chile's
  automated coding API and retained cases with high-confidence matches
  (\textgreater95\%). Treating parental occupation as a time-invariant
  origin attribute, I carried these codes back to 2016 and 2018.} is not
measured in every wave, I restrict the analysis to the 2016, 2018, and
2023 waves. After listwise deletion and restricting to key variables
(respondent occupation and the outcome measure), the final analytic
sample comprises \emph{N} = 3,435 observations nested within \emph{N} =
1,787 individuals (2016: 914; 2018: 1,377; 2023: 1,144). Consistent with
the study design, estimation proceeds within trajectory-specific
subsamples (e.g., low→low; low→middle), retaining movers and matched
non-movers from the same origin class. Then, I apply weights to
construct the corresponding counterfactual control sample for each
trajectory. Following Breen and Ermisch
(\citeproc{ref-breen_effects_2024}{2024}), I use the three waves in the
estimations that follow, allowing for correlation between the
observations for each individual in calculating the standard errors
(i.e.~clustering on the personal identifier).

@@

The outcome variable measures preferences regarding the commodification
of pensions, operationalized with a single item addressing how strongly
individuals justify conditioning access to old-age pension benefits
based on individual income. Respondents were asked: ``Is it fair in
Chile that people with higher incomes have better pensions than people
with lower incomes?'' and answered on a five-point Likert scale ranging
from 1 (``strongly disagree'') to 5 (``strongly agree''). For estimation
and interpretability, I construct a binary indicator in which values
4--5 indicate agreement with market-based access to old-age pensions
(coded 1), while values 1--3 indicate non-agreement (coded 0).
Substantively, the dependent variable is meant to capture
\emph{endorsement} of pension commodification: only responses in the
upper tail of the scale represent an active, explicit justification of
market-based differentiation, whereas neutrality corresponds to an
absence of such endorsement and is therefore treated as closer to
non-support than to support.\footnote{A potential concern is that the
  neutral category may reflect genuine ambivalence rather than
  disagreement. To address this, I conduct robustness checks using
  alternative codings: (i) excluding neutral respondents from the
  analysis; (ii) combining the neutral and agreement categories (3--5
  vs.~1--2); and (iii) estimating models with the original 5-point scale
  using linear specifications. The main mobility effects reported below
  are substantively unchanged across these specifications (see
  \hyperref[anexo]{Supplementary Material}).} This binary coding thus
yields a clear contrast between respondents who \emph{support}
market-based differentiation in pensions and those who do not. I use
this item for two main reasons: first, to enable comparisons with
existing work on market-based justice in social policy
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024}); and second,
because it taps two core dimensions of market-based welfare
distribution---(i) the centrality of economic resources as a criterion
for allocating outcomes and (ii) the framing of pensions as tradable
commodities that can be bought and sold according to ability to pay
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

I treat intergenerational occupational mobility as an exposure
indicating whether respondents occupy a different occupational status
than their fathers, closely following Breen and Ermisch's
(\citeproc{ref-breen_effects_2024}{2024}) causal framework. Occupational
assignment proceeds in two steps. First, I derive occupational status
for both origin (father) and destination (respondent) from two-digit
ISCO-08 codes using the International Socio-Economic Index of
Occupational Status (ISEI). Second, I group these ISEI scores into
terciles (low, middle, high), yielding a three-category schema for both
origin and destination strata (see Table~\ref{tbl-matrix}).
Substantively, I use ISEI because occupational status is a core
indicator of stratification and a reliable mobility measure: it locates
jobs on a hierarchical continuum defined by incumbents' typical
education level and earnings
(\citeproc{ref-hauser_intergenerational_2010}{Hauser, 2010};
\citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025}), and thus
approximates long-run socio-economic position more closely than
volatile, single-year income, which is also prone to recall and
reporting error for both parental and own resources
(\citeproc{ref-barone_rise_2022}{Barone et al., 2022}). This
occupational focus is consistent with a long tradition that interprets
the occupational structure as the backbone of the stratification system
and a key determinant of life chances
(\citeproc{ref-wrightUnderstandingClass2015}{Wright, 2015}). Moreover,
ISEI was explicitly designed to harmonize occupational stratification
across class schemes and SES measures
(\citeproc{ref-ganzeboom_internationally_1996}{Ganzeboom \& Treiman,
1996}), supporting cross-study comparability. In short, using ISEI
aligns the measurement of origin and destination on a common vertical
status continuum that is theoretically meaningful for social mobility
research and empirically feasible given the available information of
ELSOC's data.

Within each origin stratum (\(O = j\)), I estimate average treatment
effects on the treated (ATT) by comparing movers to otherwise similar
non-movers from the same origin. To reduce bias from non-random
selection into mobility based on observed characteristics, I preprocess
the data using entropy balancing, reweighting only the control group
(the immobile) so that the distribution of pre-treatment origin
covariates \(C\) among non-movers matches that of movers on selected
moments. Entropy balancing implements a maximum-entropy reweighting
scheme that chooses unit weights for non-movers to satisfy a set of
balance constraints (e.g., equality of means and, where relevant, higher
moments), while keeping the new weights as close as possible to the
original weights (\citeproc{ref-hainmueller_entropy_2012}{Hainmueller,
2012}). Substantively, this procedure addresses selection on
observables: it reweights immobile respondents so that, within each
origin class, they resemble mobile respondents in their observed
background characteristics, making any remaining differences in the
outcome more plausibly attributable to mobility rather than to
pre-existing measured advantages.

The selection of the adjustment set \(C\) follows two principles. First,
it mirrors the logic in Breen and Ermisch
(\citeproc{ref-breen_effects_2024}{2024}), who condition on attributes
determined in childhood or earlier (e.g., parental resources, household
structure, ascriptive traits). Second, it is constrained to origin-side
characteristics available in ELSOC. Guided by theory and Chilean
evidence on the determinants of relative mobility (mainly
\citeproc{ref-brunori_inequality_2025}{Brunori et al., 2025};
\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014};
\citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025};
\citeproc{ref-torche_unequal_2005}{Torche, 2005}), I include six
covariates covering family resources, household composition, and
ascriptive status: (a) parental education (highest of father/mother), in
10 ordinal categories from no schooling to postgraduate studies; (b)
co-residence with both parents at age 15 (0 = no, 1 = yes); (c)
nationality (0 = non-Chilean, 1 = Chilean); (d) age in years; (e) sex (0
= male, 1 = female); and (f) indigenous ethnicity (0 = no, 1 = yes).
This set of covariates is fixed prior to the destination class and
constitutes the maximum set on the origin side in ELSOC, thus allowing
for the plausible capture of both family and contextual influences that
may affect career trajectories, making the conditional independence
hypothesis more credible within each \(O = j\). Descriptive statistics
for all variables are presented in Supplementary Material Table A1.

After achieving balance, I construct stabilized IPW-ATT weights by
setting treated (mobile) cases to weight 1 and assigning the
entropy-balancing weights to controls, rescaled within each origin
stratum so that the mean weight remains close to one. This aligns the
counterfactual distribution of non-movers with the covariate profile of
movers in each origin class (see \hyperref[anexo]{Supplementary
Material} for balance diagnostics). The resulting weights are then used
in the outcome stage to estimate the causal effect of intergenerational
mobility on preferences for the commodification of pensions.

Because the dependent variable---the preference for pension
commodification---is binary \((Y_{it}\in{0,1})\), I estimate weighted
linear probability models (WLS) using stabilized inverse probability
weights (IPW) obtained via entropy balancing. Although OLS is often
questioned with binary outcomes, it consistently estimates the
conditional mean \(E(Y|X)\) under standard exogeneity
\(E[\varepsilon_i|X_i]=0\), and heteroskedasticity can be handled with
robust or clustered standard errors
(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009},
ch.~7.5). WLS with IPW further improves efficiency and implements the
ATT estimand by recreating the counterfactual distribution of non-movers
(\citeproc{ref-gelman_regression_2020}{Gelman et al., 2020, pp.
270--272}). Linear models are preferred here because coefficients are
directly interpretable as average differences in predicted
probabilities, whereas non-linear models complicate both weighting and
interpretation (\citeproc{ref-gelman_regression_2020}{Gelman et al.,
2020}, ch.~13).

The specification is doubly robust: entropy-balancing weights are
combined with covariate adjustment so consistency holds if either the
weighting model or the outcome model is correctly specified, while also
improving efficiency
(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). These
results are virtually identical to baseline models without covariates
(see \hyperref[anexo]{Supplementary Material} for complete models), with
no substantive changes in the mobility coefficients. Heterogeneous
effects by meritocratic perceptions and survey wave are examined through
interaction models described in the previous section.

To evaluate the robustness of the findings, I implement two sets of
sensitivity analyses. First, I conduct standard robustness checks by
re-estimating the models under alternative codings of the outcome: (i)
excluding neutral responses from the analysis, (ii) combining neutral
responses with agreement (3--5 vs.~1--2), and (iii) using the original
5-point item in a linear specification. Second, I assess sensitivity to
unmeasured confounding using the omitted-variable--bias framework of
Cinelli and Hazlett (\citeproc{ref-cinelli_making_2020}{2020}), which
quantifies how strong an unobserved confounder would need to be to
explain away the estimated ATT.

\subsection{Mobility effects models}\label{mobility-effects-models}

Table~\ref{tbl-inttime} shows the causal heterogeneity of the mobility
effect across the different waves of the survey. Regarding temporal
heterogeneity, the only statistically significant interaction term is
observed for the High→Middle trajectory in 2023
(\(\beta_{\text{int}} = 0.19\)), 95\% CI {[}0.01, 0.38{]}). However,
when this interaction is added to the baseline effect in 2016 (-0.19),
the marginal effect of High→Middle mobility in 2023 is approximately
zero, and its confidence interval includes the null. Thus, rather than
indicating a reversal toward a positive effect, the evidence suggests
that the negative impact of downward mobility from high backgrounds is
confined to 2016 and attenuates over time. No comparable temporal
changes are observed in the remaining trajectories. Overall, the results
provide little support for strong time-conditional heterogeneity in
mobility effects, beyond the fading of the initially negative
High→Middle effect.

\subsection{Robustness check and sensitivity
analysis}\label{robustness-check-and-sensitivity-analysis}

As a robustness check, I re-estimate the models under three alternative
codings of the outcome: (i) recoding the neutral category as agreement
(3--5 vs.~1--2), (ii) excluding neutral responses from the analysis, and
(iii) using the original 5-point item in a linear specification. Across
these alternative models (see \hyperref[anexo]{Supplementary Material}
for complete tables), the substantive pattern of the main effects is
preserved: the Middle→High trajectory remains positively associated with
support for pension commodification, and the High→Low trajectory remains
negatively associated with it, with both coefficients generally
increasing in magnitude. By contrast, the negative High→Middle effect
becomes statistically indistinguishable from zero in the robustness
specifications, indicating that this particular estimate is more
sensitive to how the outcome is coded than the other two trajectories.
